Bootstrapping Multi-atlas Segmentation Using Multiple Automatically Generated Templates for the Segmentation of the Whole Hippocampus and Subfields

### 2013-12-23 paper.Rnw
Discussion complete

% shortcuts 
\newcommand{\mb}{MAGeT-Brain}
\newcommand{\multiatlas}{multi-atlas}
\newcommand{\ants}{ANTS}
\newcommand{\animal}{ANIMAL}
\newcommand{\adnidataset}{ADNI1:Complete 1Yr 1.5T}
\newcommand{\FEPdataset}{SZ-FEP}
\newcommand{\fsl}{FSL FIRST}
\newcommand{\freesurfer}{FreeSurfer}
\newcommand{\maper}{MAPER}
\newcommand{\gs}{Gold Standard}
\newcommand{\silver}{silver standard}
\newcommand{\prsnr}{Pruessner}
\newcommand{\pseg}{\prsnr-protocol manual segmentation}
\newcommand{\psegs}{\prsnr-protocol manual segmentations}
\newcommand{\snt}{SNT}

@@

Monte Carlo Cross-Validation (MCCV) \citep{Shao1993} was performed using images
from the ADNI dataset \citep{Hsu2002}, and manual whole hippocampus
segmentations following the protocol laid out in \citep{Pruessner2000}.

This form of cross-validation allows us to rigorously validate a large number of
parameter settings of \mb{} (atlas and template library sizes, registration
algorithm, and label fusion method) and select the best parameters to use in
subsequent experiments.

@@

In this experiment we conduct 10 rounds of \mb{} and \multiatlas{} segmentation of
of 60 subjects using atlas and template library sizes varying from 1-9 and 1-20
images respectively, two registration algorithms (\ants{} or \animal{}), and three
label fusion techniques (unweighted, cross-correlation, and normalised mutual
information weighted majority vote).  Computed segmentations are compared to
\psegs using Dice's Similarity Coefficient (DSC).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Winterburn-XVAL: volume plots
\begin{figure}[h!]
<<WAval-vol-plots,cache=T,fig.width=7,fig.height=6>>=
wa.xval.volumes = melt(wa.xval.casted, id.vars=c('file', 'region', 'gold_0.3mm'))
wa.xval.volumes = rename(wa.xval.volumes, c("variable"="measure", "value"="volume"))
levels(wa.xval.volumes$measure) <- list("Subsampled atlas labels"="gold_0.9mm",
                               "MAGeT on subsampled atlas images" = "mb_0.9mm",
                               "MAGeT on T1 BRAVO images"="mb_bravo")

wa.xval.vols.lm = ddply(wa.xval.volumes, c("region", "measure"), function(df) {
   model = lm(volume ~ gold_0.3mm, df)
   pearson = cor(df$volume, df$gold_0.3mm, use='pairwise.complete.obs')
   data.frame(
     eqn  = lm_eqn(model), 
     corr = pearson)})
wa.xval.vols.lm$yoffset = c( 1225 - 0:2 * 40,
                             465  - 0:2 * 15, 
                             750  - 0:2 * 30,
                             325  - 0:2 * 18, 
                             1100 - 0:2 * 40 )

wa.xval.vols.lm$xoffset = c( rep(700, 3),
                             rep(250, 3),
                             rep(490, 3),
                             rep(140, 3), 
                             rep(570, 3) )

ggplot(wa.xval.volumes, aes(x=gold_0.3mm, y=volume, colour=measure)) + 
  facet_wrap(~region, ncol=3, scales='free') + 
  geom_point() + 
  coord_equal() + 
  geom_smooth(method='lm')  + 
  xlab(expression("Manual Unilateral Volume" (mm^3))) +
  ylab(expression("Automated Unilateral Volume" (mm^3))) +
  geom_text(aes(x=xoffset, y=yoffset, label=eqn), parse=TRUE,
    data=wa.xval.vols.lm, size=2,hjust=0) +
  scale_color_discrete(name="") + 
  theme(legend.direction = "horizontal", legend.position = "bottom")

@@

In this manuscript we have presented the implementation and validation of the
\mb{} framework -- a methodology that requires very few input atlases in
order to provide accurate and reliable segmentations.  Experiment 1 establishes
\mb{} as effective and as comparable to \multiatlas{} segmentation. This experiment
also rigorously characterizes the behaviour of \mb{} whole hippocampal
segmentation under various parameter settings, allowing us to choose an optimal
setting for subsequent experiments. Experiment 2 demonstrates the accuracy of
\mb{} hippocampal subfield identification despite contrast and resolution
limitations in standard T1-weighted image volumes.  Experiments 2 and 3 validate
\mb{} whole hippocampal segmentation accuracy and consistency on populations with
different ageing and neuropsychiatric characteristics.  Furthermore, taken
together, these experiments demonstrate that algorithmic performance is not
dependent on a single definition of the hippocampus but is effective with
differing hippocampal definitions \citep{Winterburn2013,Pruessner2000,Hsu2002}.

Throughout the cross validation in Experiment 1 (10-fold Monte Carlo cross
validation in the \adnidataset{} dataset subsample) we find that two parameter
choices improve segmentation accuracy: increasing the number of atlases, and the
number of templates. However, after setting the parameters to 5 atlases and 15
templates there are diminishing returns with respect to this improvement.

Previous work on \multiatlas{} segmentation methods
\citep{Aljabar2009,Collins2010} has found that cross-correlation and normalized
mutual information-based weighted label fusion improves segmentation accuracy
over simple majority vote label fusion. Selectively filtering out atlases with
lower image similarity is believed to reduce sources of error from estimating
deformations via nonlinear registration, partial volume effects from nearest
neighbour image resampling, and neuroanatomical mismatch between atlases and
subjects. That \mb{} does not see the same boost in performance from weighted
voting may suggest that the neuroanatomical variability of a template library
constructed from study subjects more closely matches any particular subject and
thereby leaving less error to filter. From our previous work on the \mb{}
algorithm we have shown that the reduction in error is not simply a smoothing or
averaging effect \citep{Chakravarty2013}.

@@

To this end, more work is required to determine the source of the slight
decrease in segmentation performance when the number of templates are set to an
even number.  Our initial concern was that this dip in performance was a
by-product of the \mb{} algorithm itself.  However, this this pattern is also
found in the results of the \multiatlas{} segmentations we used in our
experiments.  We believe that our majority voting methodology is biased towards
labels with the lowest numeric values when breaking ties (by way of the
implementation of the \verb+mode+ function used to determine majority), thus
causing the slight bias observed when using an even number of templates. This is
another area where the voting scheme could be used to improve performance.
However, it is worth noting that this limitation was previously identified by
\citet{Heckemann2006a} and, subsequently, other groups have not even considered
the potential pitfalls of an even number of candidate labels (e.g.
\citet{Leung2010}).

Another concern is the moderate-to-small improvement observed in \mb{} in
comparison to \multiatlas{} segmentation when using the same number of atlases.
The actual benefit in using \mb{} is consistency of the labelling regardless of
atlas or template choice. This is an important consideration that few have
touched on previously. The 10-fold Monte Carlo leave-one-out cross-validation
that we present in Experiment 1 is amongst one of the most stringent performed
in the \multiatlas/segmentation literature. To the best of our knowledge, other
groups using ADNI data for validation do at most a single round of
leave-one-out-validation (Table \ref{tab:other-methods}), aside from
\citep{Wang2011} who perform a similarly rigorous validation as us. The
thoroughness of our validation suggests that our results are reflective of a
true average over the choice of parameter settings and are independent of atlas
or template choice (provided the input atlases are consistently segmented).

On that note, one author (JW), an expert manual rater, identified regular
inconsistencies in the \snt{} segmentations: occurrences of over- and
under-estimation, as well as misalignments of the entire segmentation volume
(Figure \ref{fig:ADNI-segmentations}). Although the \snt{} segmentations are used
as gold standards for validation in many other studies (Table
\ref{tab:other-methods}) providing a benchmark for comparison, their use in
Experiments 1 and 4 presents the possibility that the DSC scores and
correlations we obtained are lower than what would be found with a more
accurate and consistent gold standard.

% Validation of ADNI in other works
% Chupin 2009: LOOCV, 1-fold validation
% Heckemann 2011: QC of segmentations, comparison across field strengths, 
%                 agreement between two different atlas sets
% Coupe 2011e: LOOCV
% Lotjonen 2010: LOOCV, 1-fold validation
% Leung 2010: 1-fold, QC, classification
% Wolz 2010: 1-fold over different leap settings, classification
% Collins 2010: 1-fold validation, QC
% Morra 2008: LOOCV with 21 atlases, 2 manual raters
% Wang 2011: 10 rounds of MCCV with 20 atlases and 20 subjects, chosen from 
%            a pool of 139 images

In comparison to other methodologies in the field \mb{} performs favourably. Table
\ref{tab:other-methods} surveys some of the most recent reported DSC values
reported on ADNI dataset, using \snt{} segmentations for the atlas library and as
gold standards for evaluation.  While it is difficult to compare segmentation
results across studies, gold standards, evaluation metrics, and algorithms it is
worth noting that the methods summarized require more atlases (between 16-55)
than our \mb{} implementation with the Winterburn atlases \citep{Winterburn2013}.

@@

Among the automated segmentation methods we compared in this paper
(\freesurfer{}, \maper{}, \fsl{}-FIRST), we find extremely variable
performance of all methods.  With the exception of \fsl{} all methods correlate
well with the \snt{} volumes provided in the ADNI database.  However, \freesurfer{}
and \fsl{} provide radically different definitions of the size of the hippocampus
in comparison to the other methods.  Further, when estimating bias of these
methods relative to \snt{} hippocampal volumes we see that large hippocampi are
over estimated while small hippocampi are under estimated. By comparison, \mb{}
and \maper{} are far more conservative in volume estimation, suggesting these
methods may be better suited for estimating true-positives, especially in
neurodegenerative disease subjects featuring smaller overall hippocampi.
However, in this analysis we have only compared methods by total hippocampal
volume, and so more work is needed to understand the full extent to which these
methods differ.

@@

Finally, we have also demonstrated that our algorithmic framework is
appropriate for the segmentation of hippocampal subfields in standard
T1-weighted data.  Subfield segmentation is a burgeoning topic in the
literature although very few automated methods are available for the
segmentation of 3T data \citep{Yushkevich2009,VanLeemput2009}. While recent
work demonstrates that subfield segmentations can be used for classification of
AD, MCI, and CN, there has been no explicit validation of the methodology based
on accuracy or precision.  Although the initial hippocampal subfield
segmentations from the Yushkevich group have been demonstrated to work on the
ADNI population, there has also been no validation of their work. In addition,
their work requires some manual initialization to properly function.  Our work
demonstrates that we can reliably identify the CA1, subiculum, and CA4/dentate
with only modest amounts of error.  The fact that CA2/CA3 and molecular layers
cannot be reasonably identified should not be surprising as these are extremely
thin and spatially convoluted regions that originally required high-resolution
MRI for identification.  It is likely that the extents of these regions are
well below the resolution offered by standard T1-weighted images and, in fact,
many manual segmentation methodologies do not attempt to parse these regions
either \citep{Wisse2012, Mueller2009}.

In conclusion, we have presented a flexible \multiatlas{} framework that has
considerable advantages over other methods since only a small number of atlases
is required to initialize the algorithm.  We demonstrate that our method works
robustly over hippocampal definitions, different disease populations, and
different acquisition types. Finally, we also demonstrate that accurate
identification of the hippocampal subfields is possible.

>>>>

% shortcuts 
\newcommand{\mb}{MAGeT-Brain}
\newcommand{\multiatlas}{multi-atlas}
\newcommand{\ants}{ANTS}
\newcommand{\animal}{ANIMAL}
\newcommand{\adnidataset}{ADNI1:Complete 1Yr 1.5T}
\newcommand{\FEPdataset}{SZ-FEP}
\newcommand{\fsl}{FSL FIRST}
\newcommand{\freesurfer}{FreeSurfer}
\newcommand{\maper}{MAPER}
\newcommand{\prsnr}{Pruessner}
\newcommand{\pseg}{\prsnr-protocol manual segmentation}
\newcommand{\psegs}{\prsnr-protocol manual segmentations}
\newcommand{\snt}{SNT}

@@

Monte Carlo Cross-Validation (MCCV) \citep{Shao1993} was performed using images
from the ADNI dataset \citep{Jack2008}, and manual whole hippocampus
segmentations following the protocol laid out in \citep{Pruessner2000}.  This
form of cross-validation allows us to rigorously validate a large number of
parameter settings of \mb{} (atlas and template library sizes, registration
algorithm, and label fusion method) and select the best parameters to use in
subsequent experiments.

In the Supplementary Materials we have replicated this experiment using the \snt
semi-automated labels included with the ADNI dataset.

@@

In this experiment we conduct 10 rounds of \mb{} and \multiatlas{} segmentation of
of 60 subjects using atlas and template library sizes varying from 1-9 and 1-20
images respectively, two registration algorithms (\ants{} or \animal{}), and three
label fusion techniques (unweighted, cross-correlation, and normalised mutual
information weighted majority vote).  Computed segmentations are compared to
\psegs using Dice's Similarity Coefficient (DSC).  In the Supplementary
Materials we have replicated this experiment using the \snt semi-automated
labels included with the ADNI dataset.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Winterburn-XVAL: volume plots
\begin{figure}[h!]
<<WAval-vol-plots,cache=T,fig.width=7,fig.height=6>>=
wa.xval.volumes = melt(wa.xval.casted, id.vars=c('file', 'region', 'gold_0.3mm'))
wa.xval.volumes = rename(wa.xval.volumes, c("variable"="measure", "value"="volume"))
levels(wa.xval.volumes$measure) <- list("Subsampled atlas labels"="gold_0.9mm",
                               "MAGeT on subsampled atlas images" = "mb_0.9mm",
                               "MAGeT on T1 BRAVO images"="mb_bravo")
wa.xval.vols.lm = ddply(wa.xval.volumes, c("region", "measure"), function(df) {
   model = lm(volume ~ gold_0.3mm, df)
   pearson = cor(df$volume, df$gold_0.3mm, use='pairwise.complete.obs')
   data.frame(
     eqn  = lm_eqn(model), 
     corr = pearson)})
wa.xval.vols.lm$yoffset = c( 1225 - 0:2 * 40,
                             465  - 0:2 * 15, 
                             750  - 0:2 * 30,
                             325  - 0:2 * 18, 
                             1100 - 0:2 * 40 )
wa.xval.vols.lm$xoffset = c( rep(700, 3),
                             rep(250, 3),
                             rep(490, 3),
                             rep(140, 3), 
                             rep(570, 3) )
ggplot(wa.xval.volumes, aes(x=gold_0.3mm, y=volume, colour=measure)) + 
  facet_wrap(~region, ncol=3, scales='free') + 
  geom_point() + 
  coord_equal() + 
  geom_smooth(method='lm')  + 
  xlab(expression("Manual Unilateral Volume" (mm^3))) +
  ylab(expression("Automated Unilateral Volume" (mm^3))) +
  geom_text(aes(x=xoffset, y=yoffset, label=eqn), parse=TRUE,
    data=wa.xval.vols.lm, size=2,hjust=0) +
  scale_color_discrete(name="") + 
  theme(legend.direction = "horizontal", legend.position = "bottom")

@@

In this manuscript we have presented the implementation and validation of the
\mb{} framework -- a methodology that requires very few input atlases in order
to provide accurate and reliable segmentations.  Experiment 1 compares \mb{} to
basic-\multiatlas{} segmentation by characterising the behaviour of each
approach with varying parameter settings, and allows us to choose an optimal
setting for subsequent experiments. Experiment 2 demonstrates the reliability of
\mb{} in producing subfield segmentations which match the segmentation protocol
of the input atlases despite contrast and resolution limitations in standard
T1-weighted image volumes. Experiments 2 and 3 validate whole hippocampal
segmentation precision and consistency on populations with different ageing and
neuropsychiatric characteristics. All of these experiments together demonstrate
that \mb{} algorithmic performance is not dependent on a single definition of
the hippocampus but is effective with differing hippocampal definitions
\citep{Winterburn2013,Pruessner2000,Hsu2002}.

The core claim the \mb{} method is based on, that we can meaningfully bootstrap
a template library from a small set of labelled atlas images, is validated in
the cross validation conducted in Experiment 1.  We find that both increasing
the number of atlases and the number of templates used improves \mb{}
segmentation over and above basic-\multiatlas segmentations using the same
number of atlas images. That is, by taking the extra step of generating a
template library from a small atlas set \mb{} is able to improve the final
segmentation similarity. The magnitude of this improvement is greatest with a
small number of atlases, but even with larger atlas libraries we have found that
generating a template library reduces the variability in segmentations (i.e.
\mb{} more consistently produces high quality segmentations than does
basic-\multiatlas).

Interestingly, previous work on \multiatlas{} segmentation methods
\citep{Aljabar2009,Collins2010} has found that cross-correlation and normalized
mutual information-based weighted label fusion improves segmentation accuracy
over simple majority vote label fusion, and yet we did not see a significant
indication of this effect in the \mb{} segmentations. Selectively filtering out
atlases with lower image similarity is believed to reduce sources of error from
estimating deformations via nonlinear registration, partial volume effects from
nearest neighbour image resampling, and neuroanatomical mismatch between atlases
and subjects. That \mb{} does not see the same boost in performance from
weighted voting may suggest that the neuroanatomical variability of a template
library constructed from study subjects more closely matches any particular
subject and thereby leaving less error to filter. From our previous work on the
\mb{} algorithm we have shown that the reduction in error is not simply a
smoothing or averaging effect \citep{Chakravarty2013}.

@@

More work is required to determine the source of the slight decrease in
segmentation performance when the number of templates are set to an even number.
Our initial concern was that this dip in performance was a by-product of the
\mb{} algorithm itself.  However, this this pattern is also found in the results
of the \multiatlas{} segmentations we used in our experiments.  We believe that
our majority voting methodology is biased towards labels with the lowest numeric
values when breaking ties (by way of the implementation of the \verb+mode+
function used to determine majority), thus causing the slight bias observed when
using an even number of templates. This is another area where the voting scheme
could be used to improve performance.  However, it is worth noting that this
limitation was previously identified by \citet{Heckemann2006a} and,
subsequently, other groups have not even considered the potential pitfalls of an
even number of candidate labels (e.g.  \citet{Leung2010}).

Another concern is the moderate improvement observed in \mb{} in comparison to
\multiatlas{} segmentation when using the same number of atlases.  The actual
benefit in using \mb{} is consistency of the labelling regardless of atlas or
template choice, as mentioned above. This is an important consideration that few
have touched on previously. The 10-fold Monte Carlo cross-validation that we
present in Experiment 1 is amongst one of the most stringent performed in the
\multiatlas/segmentation literature. To the best of our knowledge, with the
exception of \citep{Wang2011}, other groups using ADNI data for validation do at
most a single round of leave-one-out-validation (Table \ref{tab:other-methods}).
The thoroughness of our validation suggests that our results are reflective of a
true average over the choice of parameter settings and are independent of atlas
or template choice (provided the input atlases are consistently segmented).

On that note, one author (JW), an expert manual rater, identified regular
inconsistencies in the \snt{} segmentations: occurrences of over- and
under-estimation, as well as misalignments of the entire segmentation volume
(Figure \ref{fig:ADNI-segmentations}). Although the \snt{} segmentations are
used as benchmarks for validation in many other studies (Table
\ref{tab:other-methods}), these segmentation inconsistencies present the
possibility that a more accurate and consistent benchmark segmentation protocol
ought to be used in order to truly understand the results of such validations.
Indeed, our replication of the 10-fold cross-validation using \snt{}
segmentations (Experiment 5) shows noticeably poorer mean similarity scores for
both \mb{} and \multiatlas.

Thus, in comparison to other methodologies in the field \mb{} performs
favourably. Table \ref{tab:other-methods} surveys some of the most recent
reported DSC values reported on ADNI dataset, using \snt{} segmentations for the
atlas library and as gold standards for evaluation.  While it is difficult to
compare segmentation results across studies, gold standards, evaluation metrics,
and algorithms it is worth noting that the methods summarized require more
atlases (between 16-55) than our \mb{} implementation with the Winterburn
atlases \citep{Winterburn2013}.

@@

Among the automated segmentation methods we compared in this paper
(\freesurfer{}, \maper{}, \fsl{}), we find extremely variable
performance of all methods.  With the exception of \fsl{} all methods correlate
well with the \snt{} volumes provided in the ADNI database.  However, \freesurfer{}
and \fsl{} provide radically different definitions of the size of the hippocampus
in comparison to the other methods.  Further, when estimating bias of these
methods relative to \snt{} hippocampal volumes we see that large hippocampi are
over estimated while small hippocampi are under estimated. By comparison, \mb{}
and \maper{} are far more conservative in volume estimation, suggesting these
methods may be better suited for estimating true-positives, especially in
neurodegenerative disease subjects featuring smaller overall hippocampi.
However, in this analysis we have only compared methods by total hippocampal
volume, and so more work is needed to understand the full extent to which these
methods differ.

@@

Finally, we have provide evidence that using the Winterburn high-resolution
hippocampal subfield atlases \citep{Winterburn2013} our algorithmic framework is
appropriate for the segmentation of hippocampal subfields in standard
T1-weighted data. Subfield segmentation is a burgeoning topic in the literature
although very few automated methods are available for the segmentation of 3T
data \citep{Yushkevich2009,VanLeemput2009}. While recent work demonstrates that
subfield segmentations can be used for classification of AD, MCI, and CN, there
has been no explicit validation of the segmentations being produced other than
indirectly via classification accuracy (for instance, the semi-automated
subfield segmentations produced in \citet{Yushkevich2009}).  Our work
demonstrates that we can reliably produce CA1, subiculum, and CA4/dentate
subfield segmentations that correspond to those in the input atlas with only 
only modest amounts of error in volume.  The fact that CA2/CA3 and molecular layers
cannot be reasonably reproduced should not be surprising as these are extremely
thin and spatially convoluted regions that originally required high-resolution
MRI for identification and so it is likely that the extents of these regions are
well below the resolution and contrast offered by standard T1-weighted images.
In addition, some manual subfield segmentation methodologies do not attempt to parse
these regions even on high-resolution scans (\citep{Wisse2012, Mueller2009}; see
also Table \label{tab:subfield-segmentation-protocols}). Note, it is because of
both these reasons (the inability to discern subregions in standard T1-weighted
images, and the definitional uncertainty in subfield segmentation at the present
moment) that we opted to speak of the precision or reliability of \mb{} subfield
segmentation rather than speak of accuracy.

In conclusion, we have presented a flexible \multiatlas{}-based framework that has
considerable advantages over other methods since only a small number of atlases
is required to initialize the algorithm.  We demonstrate that our method works
robustly over hippocampal definitions, different disease populations, and
different acquisition types. Finally, we also demonstrate that accurate
identification of the hippocampal subfields is possible.
